---
output:
  pdf_document: default
  html_document: default
---

------------------------------------------------------------------------

# Data for: Is the short-term phenology and long-term trajectory of river productivity changing over time?

## Nick Marzolf, 2024-01-29

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

RMarkdown file to accompany: Is the short-term phenology and long-term trajectory of river productivity changing over time?, a manuscript submitted to Limnology and Oceanography: Letters Special Issue: "Changing Phenology in Aquatic Ecosystems".

Nicholas S. Marzolf^1,\*^, Michael J. Vlah^1^, Heili E. Lowman^2^, Weston M. Slaughter^1^, Emily S. Bernhardt^1^

Affiliations

^1^ Department of Biology, Duke University, Durham, NC, USA

^2^ Department of Natural Resources and Environmental Science, University of Nevada, Reno, Reno, Nevada, USA

\*corresponding author: [nick.marzolf\@jonesctr.org](mailto:nick.marzolf@jonesctr.org){.email} (ORCiD: 0000-0001-9146-1643)

To recreate this HTML, load the data files into a folder named 'data' at the same path as the .rmd file, open the .rmd file, and click 'Knit'.

# Load packages

```{r load-packages, echo = TRUE, message = FALSE, warning = FALSE}
# data manipulation
library(tidyverse)
library(dplyr)
library(lubridate)
library(purrr)
library(tibble)
library(broom)

# plotting
library(ggplot2)
library(ggpubr)
library(ggExtra)
library(GGally)
library(gghighlight)
source("code/theme_nick.R")
ggplot2::theme_set(theme_nick())

# spatial
library(sf)
library(spData)

# statistics
library(lme4)
library(nlme)
library(performance)
library(EnvStats)
library(EflowStats)
library(car)
library(trend)
library(merTools)

# reproducability
library(pander)
```

```{r define-constants, echo=TRUE, message=FALSE, warning=FALSE}
# define constants and axis labels
metab_units_area <- (expression(paste('GPP (g C ', m^-2, ' ',y^-1,')', sep = ' ')))

low_p <- '#103801'
high_p <- '#43fa00'

# quotient to convert GPP from O2 to C
# g C = g O2 * (1 mol O2/32 g O2) * (2 mol O/1 mol O2) * (1 mol C/1 mol O) * (12 g C/1 mol C)
rq <- (1/32)*(2/1)*(1/1)*(12/1) 

# where to save figures
save_figs <- FALSE

dpi <- 1200

version <- 'v3'

fig_dir <- glue::glue('manuscript/long-term_GPP/figures/{version}/')

if(!dir.exists(fig_dir)){
  dir.create(fig_dir)
}
```

# Explore sites- MOVE ME DOWN

```{r sites, warning = FALSE, echo = TRUE}
# histogram of watershed areas
table_s1 <- readr::read_csv('data/data_citation/1_site_info.csv')
table_s1

readr::write_csv(table_s1,
                 'data/data_citation/ms_usgs_annual_metab/table_s1_site_info.csv')

table_s1_plot <- table_s1 %>% 
  sf::st_as_sf(.,
               coords = c('Longitude', 'Latitude'),
               crs = 4326)

fig_s1_map <- ggplot()+
  geom_sf(data = world %>% 
            dplyr::filter(name_long == 'United States') %>% 
            sf::st_set_crs(4326),
          fill = 'white')+
  geom_sf(data = table_s1_plot)+
  # geom_sf_label(data = table_s1_plot,
  #              aes(label = `Clean Name`),
  #              check_overlap = TRUE)+
  coord_sf(xlim = c(-130, -65), ylim = c(23, 55), expand = FALSE)


fig_s1_hist <- ggplot(table_s1,
                      aes(x = `Drainage Area (km2)`*2.58999))+ # convert to km2
  geom_histogram()+
  labs(x = expression(paste('Watershed area (', km^2,')')),
       y = 'Count')+
  scale_x_log10()
fig_s1_hist
```

```{r fig-s1, fig.width = 9, fig.height = 4}
fig_s1 <- cowplot::plot_grid(fig_s1_map,
                             fig_s1_hist,
                             ncol = 2, 
                             rel_widths = c(2,1),
                             axis = 'b',
                             align = 'v')
fig_s1

if(save_figs){
  ggsave(plot = fig_s1,
         glue::glue(fig_dir, 'fig_s1.png'),
         dpi = dpi,
         width = 8.5, height = 3)
}
```

# Load GPP data

```{r load-data, warning = FALSE, echo = TRUE}

# modified function from 'code/metabolism_synthesis/Part II- Filtering and gap filling.Rmd
source('code/functions/fill_and_normalize_metab.R')

river_metab <- readr::read_csv('data/data_citation/5_usgs_metabolism.csv') 
river_metab_filled <- fill_and_normalize_metab(river_metab)

head(river_metab_filled)

n_obs <- length(river_metab_filled$GPP_filled)

n_sites <- length(unique(river_metab_filled$site))

```

# 1) Calculate cumulative daily GPP and total annual GPP

```{r GPP-cumsum, warning = FALSE, echo = TRUE}

# calculate site-year cumulative GPP
river_metab_cumsum <- river_metab_filled %>% 
  dplyr::mutate(year = lubridate::year(date)) %>% 
  dplyr::group_by(site, year) %>% 
  dplyr::mutate(jday = lubridate::yday(date),
                GPP_C_cumsum = cumsum(GPP_filled*rq)) 

# calculate annual GPP for each site-year
metab_all_annual <- river_metab_filled %>% 
  dplyr::mutate(year = lubridate::year(date)) %>% 
  dplyr::group_by(site, year) %>% 
  dplyr::summarise(GPP_daily_mean = mean(GPP_filled*rq, na.rm = TRUE),
                   GPP_ann_sum = sum(GPP_filled*rq, na.rm = TRUE),
                   GPP_daily_cv = EnvStats::cv(GPP_filled*rq, 
                                               na.rm = TRUE, 
                                               method = 'l.moments')*100,
                   Q_daily_mean  = mean(discharge, na.rm = TRUE),
                   Q_daily_cv = EnvStats::cv(discharge,
                                             na.rm = TRUE,
                                             method = 'l.moments')*100,
                   temp_daily_mean = mean(temp.water, na.rm = TRUE),
                   temp_daily_cv = EnvStats::cv(temp.water,
                                                na.rm = TRUE,
                                                method = 'l.moments')*100) %>% 
  dplyr::mutate(clean_site_num = as.numeric(substr(site, 6,30))) %>% 
  dplyr::left_join(table_s1 %>% 
                     rename(clean_site_num = `Site Number`) , 
                   'clean_site_num')
summary(metab_all_annual$GPP_ann_C)


# plot of annual productivity time-series
fig_s2 <- ggplot2::ggplot(metab_all_annual,
                          aes(x = year,
                              y = GPP_ann_sum,
                              group = site,
                              color = GPP_ann_sum))+
  ggplot2::geom_line()+
  ggplot2::geom_point()+
  ggplot2::scale_color_gradient(name = metab_units_area,
                                low = low_p,
                                high = high_p)+
  ggplot2::ylab(metab_units_area)+
  ggplot2::theme(axis.title.x = element_blank(),
                 legend.position = 'none')+
  scale_y_continuous(limits = c(0 , 4000))
fig_s2

if(save_figs){
  ggsave(plot = fig_s2,
         glue::glue(fig_dir,'fig_s2.png'),
         dpi = dpi, 
         width = 6, height = 4)
}



# calculate median annual and CV for each site time series
site_mean_ann_gpp <- metab_all_annual %>% 
  dplyr::group_by(site) %>% 
  dplyr::summarise(median_ann_GPP = median(GPP_ann_sum, na.rm = TRUE),
                   sd_ann_GPP = sd(GPP_ann_sum, na.rm = TRUE),
                   min_ann_GPP = min(GPP_ann_sum, na.rm = TRUE),
                   max_ann_GPP = max(GPP_ann_sum, na.rm = TRUE),
                   GPP_percent_diff = ((max_ann_GPP - min_ann_GPP)/min_ann_GPP)*100,
                   GPP_ann_cv = EnvStats::cv(GPP_ann_sum, na.rm = TRUE, method = 'l.moments')*100,
                   mean_ann_temp = mean(temp_daily_mean, na.rm = TRUE),
                   mean_sd_temp = sd(temp_daily_mean, na.rm = TRUE),
                   cv_mean_temp = EnvStats::cv(temp_daily_mean, method = 'l.moments')*100,
                   median_ann_discharge = median(Q_daily_mean, na.rm = TRUE),
                   median_sd_discharge = sd(Q_daily_mean, na.rm = TRUE),
                   cv_median_discharge = EnvStats::cv(Q_daily_mean, na.rm = TRUE, method = 'l.moments')*100) %>% 
  dplyr::arrange(desc(max_ann_GPP)) %>% 
  dplyr::mutate(clean_site_num = as.numeric(substr(site, 6,30))) %>% 
  dplyr::left_join(table_s1 %>% 
                     rename(clean_site_num = `Site Number`) , 
                   'clean_site_num')

```

```{r fig-1, fig.width=15, fig.height=15}
# cumulative site-year GPP
river_metab_cumsum_names <- river_metab_cumsum %>% 
  dplyr::mutate(clean_site_num = as.numeric(substr(site, 6,30))) %>% 
  dplyr::left_join(table_s1 %>% 
                     rename(clean_site_num = `Site Number`) , 
                   'clean_site_num')


river_metab_cumsum_names$Name <- forcats::fct_relevel(river_metab_cumsum_names$Name,
                                                      site_mean_ann_gpp %>% 
                                                        dplyr::pull(`Clean Name`)
                                                      )

y_axis_title <- bquote(Cumulative ~ .(metab_units_area))

fig_1 <- ggplot(river_metab_cumsum_names, 
                aes(x = jday,
                    y = GPP_C_cumsum,
                    group = interaction(year, site),
                    color = GPP_C_cumsum))+
  geom_line(show.legend = FALSE)+
  facet_wrap(forcats::fct_relevel(`Clean Name`,
                                  site_mean_ann_gpp %>% 
                                    dplyr::pull(`Clean Name`)) ~ .,
             scales = 'free_y',
             ncol = 5)+
  scale_color_gradient(low = low_p,
                       high = high_p)+
  scale_y_continuous(breaks = function(y) {
    max_value <- max(y, na.rm = TRUE)
    tick_marks <- seq(0, round(max_value, digits = -2), length.out = 3)
    return(tick_marks)
  }
  )+
  labs(x = 'Day of Year',
       y = y_axis_title)+
  theme(axis.text.x = element_blank(),
        strip.text = element_text(size = 7),
        strip.background = element_rect(fill = 'white'))
fig_1

if(save_figs){
  ggsave(plot = fig_1,
         glue::glue(fig_dir, 'fig_1.png'),
         dpi = dpi, 
         width = 15, height = 13)
}

```

# 2) Annual productivity range

```{r viz-ann-prod, warning = FALSE, echo = TRUE, fig.width=11, fig.height=5}

cv <- site_mean_ann_gpp %>% 
  dplyr::select(site, GPP_ann_cv) %>% 
  left_join(.,metab_all_annual, 'site')

cv_sum <- cv %>% 
  group_by(site) %>% 
  summarise(mean_ann_GPP_C = mean(GPP_ann_sum, na.rm = TRUE), 
            sd_ann_GPP_C = sd(GPP_ann_sum, na.rm = TRUE),
            median_daily_CV = median(GPP_daily_cv, na.rm = TRUE),
            min_daily_CV = min(GPP_daily_cv), 
            max_daily_CV = max(GPP_daily_cv),
            GPP_ann_cv = first(GPP_ann_cv),
            GPP_range = max(GPP_ann_sum) - min(GPP_ann_sum),
            CV_ratio = median_daily_CV/GPP_ann_cv)

# boxplot of annual GPP for each site
fig_2_a <- ggplot(metab_all_annual %>% 
                    dplyr::mutate(clean_site_num = as.numeric(substr(site, 6,30))) %>% 
                    dplyr::left_join(table_s1 %>% 
                     rename(clean_site_num = `Site Number`) , 
                   'clean_site_num'),
                  aes(x = forcats::fct_relevel(`Clean Name`,
                                               site_mean_ann_gpp %>% 
                                                 dplyr::pull(`Clean Name`)), 
                      y = GPP_ann_sum))+
  geom_boxplot(width = 0.75,
               outlier.shape = NA,
               show.legend = FALSE,)+
  geom_point(aes(color = GPP_ann_sum),
             size = 1)+
  coord_flip()+
  scale_y_continuous(name = metab_units_area,
                     limits = c(-1, 4000))+
  scale_color_gradient(name = metab_units_area,
                       low = low_p,
                       high = high_p)+
  theme(axis.text.y = element_text(angle = 0, size = 8,
                                   hjust = 0.95,vjust = 0.2),
        axis.title.y = element_blank(),
        legend.position = c(0.9,0.9),
        legend.justification = c(0.9,0.9),
        legend.background = element_rect(fill = NA, 
                                         color = 'black'))
fig_2_a


fig_2_b <- ggplot2::ggplot(data = site_mean_ann_gpp,
                           aes(y = median_ann_GPP, 
                               x = GPP_ann_cv,
                               group = site,
                               color = median_ann_GPP))+
  ggplot2::geom_point(size = 2)+
  geom_errorbar(aes(ymin = median_ann_GPP - sd_ann_GPP,
                    ymax = median_ann_GPP + sd_ann_GPP))+
  ggplot2::labs(x = expression(paste(CV[GPP-Inter],' (%)')),
                y = expression(paste('Median Annual GPP (g C  ',m^-2, ' ',y^-1,')')))+
  lims(y = c(0, 3000),
       x = c(5, 40))+
  scale_color_gradient(low = low_p,
                       high = high_p)+
  theme(legend.position = 'none')
fig_2_b


fig_2_c <- ggplot(cv,
       aes(x = GPP_ann_cv,
           y = GPP_daily_cv,
           color = GPP_ann_sum,
           group = site))+
  geom_point()+
  geom_line()+
  geom_smooth(method = 'lm')+
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed')+
  scale_color_gradient(name = metab_units_area,
                       low = low_p,
                       high = high_p,)+
  lims(x = c(5,40),
       y = c(20,160))+
  labs(x = expression(paste(CV[GPP-Inter], ' (%)')),
       y = expression(paste(CV[GPP-Intra], ' (%)')))+
  theme(legend.position = 'none')



fig_2 <- ggpubr::ggarrange(fig_2_a,NA,
                           ggpubr::ggarrange(fig_2_b,
                                             fig_2_c, 
                                             nrow = 2,
                                             labels = c('b', 'c'),
                                             align = 'v'),
                           widths = c(1.5,0,1),
                           labels = 'a',
                           # hjust = c(-8.5,-8),
                           # vjust = c(1.7, 1.7),
                           ncol = 3,
                           align = 'v')
fig_2

if(save_figs){
  ggsave(plot = fig_2,
         glue::glue(fig_dir, 'fig_2.png'),
         dpi = dpi,
         width = 11, height = 8)
}
```

# 3) When are rivers productive

```{r gpp-quantiles, warning = FALSE, echo = TRUE, fig.width=6, fig.height=10}
list_quants <- list()
quantiles <- c(0.25, 0.5, 0.75, 0.95)
for(i in 1:length(quantiles)){
  
  quantile <- quantiles[i]
  
  list_quants[[i]] <- river_metab_cumsum %>% 
    dplyr::select(site, date, year, GPP_C_cumsum) %>% 
    dplyr::group_by(site, 
                    year = lubridate::year(date)) %>% 
    dplyr::mutate(jday = lubridate::yday(date),
                  gpp_cdf = ecdf(GPP_C_cumsum)(GPP_C_cumsum)) %>% 
    dplyr::arrange(site, year, gpp_cdf) %>% 
    dplyr::slice(which.min(abs(gpp_cdf - quantile))) 
}
names(list_quants) <- quantiles
df_quants <- map_df(list_quants, 
                    ~as.data.frame(.x), 
                    .id = 'quantile')

# by quantile plots 

# rank sites by increasing variability of 50th percentile
quant_50_rank <- list_quants[[1]] %>% 
  dplyr::group_by(site) %>% 
  dplyr::summarise(iqr = IQR(jday)) %>% 
  dplyr::arrange(desc(iqr)) %>% 
  dplyr::pull(site)


quant_75_rank <- list_quants[[2]] %>% 
  dplyr::group_by(site) %>% 
  dplyr::summarise(iqr = IQR(jday)) %>% 
  dplyr::arrange(desc(iqr)) %>% 
  dplyr::pull(site)


quant_95_rank <- list_quants[[3]] %>% 
  dplyr::group_by(site) %>% 
  dplyr::summarise(iqr = IQR(jday)) %>% 
  dplyr::arrange(desc(iqr)) %>% 
  dplyr::pull(site)

```

```{r quant-trends, echo=TRUE, warning = FALSE, fig.height=7, fig.width=7}
# trend analysis for quantiles
quantile_trends <- df_quants %>% 
  dplyr::group_by(site, quantile) %>% 
  tidyr::nest() %>% 
  dplyr::mutate(mk_test = purrr::map(data, ~trend::mk.test(x = .$jday))) %>% 
  dplyr::mutate(mk_p = purrr::map_dbl(mk_test, ~.$p.value),
                mk_s = purrr::map_dbl(mk_test, ~.$estimates['S']),
                mk_sig = ifelse(mk_p <= 0.05, 'significant', 'non-significant'),
                mk_slope = ifelse(mk_s > 0, 'increasing', 'decreasing')) 

quantile_sum_sigs <- quantile_trends %>% 
  dplyr::ungroup() %>% 
  dplyr::select(quantile, site, 
                mk_p, mk_s, mk_sig, mk_slope)

quants_n_sig <- quantile_sum_sigs %>% 
  dplyr::filter(mk_p <= 0.05) %>% 
  dplyr::group_by(quantile,mk_slope) %>% 
  summarise(n = n()) %>% 
  data.frame()

quantile_trends_sum <- df_quants %>% 
  dplyr::left_join(quantile_sum_sigs) %>% 
  dplyr::mutate(mk_meaning = case_when(mk_slope == 'increasing' ~ 'Later',
                                       mk_slope == 'decreasing' ~ 'Earlier'))

table_s2 <- quantile_trends_sum %>% 
  dplyr::filter(mk_p < 0.05) %>% 
  dplyr::select(quantile, site, year, date, jday, mk_s,mk_p,mk_meaning) %>% 
  dplyr::mutate(across(where(is.numeric), round, 3))
readr::write_csv(table_s2,
                 'data/data_citation/ms_usgs_annual_metab/table_s2_quantile_trends.csv')

fig_3 <- ggplot(quantile_trends_sum,
                aes(x = year,
                    y = jday,
                    group = site,
                    color = mk_meaning))+
  # geom_point()+
  geom_line(linewidth = 0.6)+
  gghighlight::gghighlight(mk_sig == 'significant',
                           use_direct_label = FALSE,
                           keep_scales = TRUE,
                           unhighlighted_params = list(alpha("grey10", 0.4),
                                                       linewidth = 0.2),
                           calculate_per_facet = TRUE)+
  # ggside::geom_ysidedensity(aes(x = stat(density)))+
  scale_y_continuous(limits = c(0, 366))+
  scale_color_manual(name = element_blank(),
                     values = c(high_p, low_p))+
  facet_wrap(. ~ quantile,
             scales = 'free', 
             nrow = 2, ncol = 2)+
  labs(x = 'Year',
       y = 'Day of Year')+
  theme(legend.title = element_blank())
fig_3

if(save_figs){
  ggsave(plot = fig_3,
         glue::glue(fig_dir, 'fig_3.png'),
         dpi = dpi,
         width = 6.5, height = 5)
}

```

# 4) Annual productivity trends

```{r prod-trends, warning = FALSE, echo = TRUE, fig.width=5, fig.height=5}

annual_C_sum_trends <- metab_all_annual %>% 
  # dplyr::select(site, year, GPP_ann_C) %>% 
  dplyr::group_by(site, Name) %>% 
  tidyr::nest() %>% 
  dplyr::mutate(mk_test = purrr::map(data, ~trend::mk.test(x = .$GPP_ann_sum))) %>% 
  dplyr::mutate(mk_p = purrr::map_dbl(mk_test, ~.$p.value),
                mk_s = purrr::map_dbl(mk_test, ~.$estimates['S']),
                mk_sig = ifelse(mk_p <= 0.05, 'significant', 'non-significant'),
                mk_slope = ifelse(mk_s > 0, 'increasing', 'decreasing')) 


annual_C_sum_sigs <- annual_C_sum_trends %>% 
  dplyr::ungroup() %>% 
  dplyr::select(site, 
                mk_p, mk_s, mk_sig, mk_slope)


n_sig <- annual_C_sum_sigs %>% 
  dplyr::filter(mk_p <= 0.05) %>% 
  dplyr::group_by(mk_slope) %>% 
  summarise(n = n()) %>% 
  data.frame()

lt_changers <- annual_C_sum_sigs %>% 
  dplyr::filter(mk_p <= 0.05) %>% 
  dplyr::pull(site)

river_trends <- metab_all_annual %>% 
  dplyr::left_join(annual_C_sum_sigs) 


fig_4 <- ggplot2::ggplot(river_trends,
                         ggplot2::aes(x = year, 
                                      y = GPP_ann_sum,
                                      color = mk_s))+
  ggplot2::geom_line(aes(group = `Clean Name`),
                     linewidth = 0.6)+
  scale_y_continuous(limits = c(0, 4000))+
  scale_x_continuous(n.breaks = 10)+
  gghighlight::gghighlight(mk_sig == 'significant',
                           use_direct_label = TRUE,
                           keep_scales = TRUE,
                           unhighlighted_params = list(alpha("grey40", 0.4),
                                                       linewidth = 0.4))+
  ggplot2::labs(y = metab_units_area)+
  scale_color_gradient(name = element_blank(),
                     low = low_p, 
                     high = high_p)+
  theme(axis.title.x = element_blank(),
        legend.position = 'none',
        legend.text = element_text(size = 10),
        legend.background = element_rect(color = 'black',
                                         linewidth = 0.25),
        legend.margin = margin(-0.5, 1, 0.4, 0.1))
fig_4

if(save_figs){
  ggsave(plot = fig_4,
         glue::glue(fig_dir, 'fig_4.png'),
         dpi = dpi,
         width = 5, height = 4)
}


table_s3 <- annual_C_sum_sigs %>% 
  dplyr::select(site, mk_s,mk_p)
colnames(table_s3) <- c('Site',
                        "Sen's Slope",
                        "Sen's p-value",
                        "Mann-Kendall S",
                        "Mann-Kendall p-value")
readr::write_csv(table_s3,
                 'data/data_citation/ms_usgs_annual_metab/table_s3_trend_stats.csv')

```

# Session Info

```{r session-info}
pander::pander(sessionInfo())
```
